---
title: "Quantiles"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
  html_document:
    toc: yes
  word_document: default
geometry: margin=1in
graphics: yes
header-includes:
- \usepackage{graphicx}
- \usepackage{color}
- \usepackage{hyperref}
- \newcommand{\ve}[1]{\mathbf{#1}}    
- \newcommand{\sv}[1]{\boldsymbol{#1}} 
- \newcommand{\mat}[1]{\mathbf{#1}}    
- \newcommand{\sm}[1]{\boldsymbol{#1}}   
- \newcommand{\tr}[1]{{#1}^{\mkern-1.5mu\mathsf{T}}}    
- \newcommand{\norm}[1]{||{#1}||}           
- \newcommand{\given}{~\vline~}
- \newcommand{\indep}{\bot\hspace{-.6em}\bot}
- \newcommand{\notindep}{\bot\hspace{-.6em}\bot\hspace{-0.75em}/\hspace{.4em}}
- \newcommand{\depend}{\Join}
- \newcommand{\notdepend}{\Join\hspace{-0.9 em}/\hspace{.4em}}
- \newcommand{\imply}{\Longrightarrow}
- \newcommand{\notimply}{\Longrightarrow \hspace{-1.5em}/ \hspace{0.8em}}
- \newcommand{\code}[1]{\texttt{#1}}
- \newcommand*{\Rnsp}{\textsf{R}}
- \newcommand*{\R}{\textsf{R}$~$}
- \newcommand{\bigwig}[1]{\widetilde{#1}}

fontsize: 9pt
  
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(12314159)

imageDirectory <- "./img"
dataDirectory <- "./data"
path_concat <- function(path1, path2, sep="/") paste(path1, path2, sep = sep)
```

---

**29 marks**

1. Suppose we have a continuous random variable $X$ with distribution function $F_X(x) = Pr(X \le x)$ and quantile function
$Q_X(p) = F_X^{-1}(p)$.  That is $p = F_X(x) = Pr(X \le x)$ and $p = Pr(X \le Q_X(p)) = F_X(Q_X(p)) = F_X(F_X^{-1}(p)) = p$.

    a. *(4 marks)* Suppose $Y = aX +b$ for some constants $a > 0$ and $b$. 
        **Prove** that a plot of the parametric curve $(Q_X(p), Q_Y(p))$ for $p \in (0,1)$ must follow a straight line.  
        
        Give the equation of that line.
        $$
        p = Pr(Y \le Q_Y(p)) = Pr(aX + b \le Q_Y(p) ) = F_X(\frac{Q_Y(p) - b}{a})\\
        P = F_X(Q_X(p))
        $$
  
        Therefore
  
        $$\frac{Q_Y(p)-b}{a} = Q_X(p)\\
        Q_Y(p) = aQ_X(p) + b
        $$
        
    b. *(3 marks)* When $F_X(x)$ and $Q_X(p)$ are the cumulative distribution and quantile functions of the continuous random variable $X$, show that if $U \sim U(0,1)$, then 
    
        \[Pr(Q_X(U) \le x )  = F_X(x). \]
        We have $U \sim U(0,1)$ so that $Pr(U < a) = a$ for $a \in [0,1]$
        therefore 
        $$
        Pr(Q_X(U) \le x) = Pr(F_X(Q_X(U)) \le F_X(x)) \\
        = Pr(U \le F_X(x)) = F_X(x)
        $$

    
    c. The above result implies that we could generate $n$ independently and identically distributed (i.i.d.) random realizations $X$ from $F_X(x)$ by generating $n$ i.i.d. random realizations $U$ from $U(0,1)$ and defining $X = Q_X(U)$.  
    
        In \R the function `runif()` will generate uniform pseudo-random numbers. 
    
        (Similarly, `dunif()`, `punif()`, and `qunif()` will return the density, the distribution, and the quantile functions, respectively, for a uniform random variable.  See `help("runif")` for details.
    
        i. *(1 mark)* Write an \R function
    
            ```{r r_unigenFx, eval = TRUE}
            r_unifgenFx <- function(n, qfunction = qnorm) {
                                # Insert your code here
                                random <- runif(n)
                                qfunction(random)
                                 }
            ```
        
            which will generate and return `n` pseudo random observations from the distrubution whose quantile function is the value of the argument `qfunction`.  Show your code.
            
        ii. *(2 marks)*  Execute the following code snippets to illustrate your code
        
            ```{r hists, eval = TRUE}
            # make sure we all get the same result
            set.seed(1234567)
            # save the currentgraphical parameters and set `mfrow`
            oldPar <- par(mfrow = c(1,2))
            
            hist(r_unifgenFx(1000),main = "normal", xlab="p", col = 'maroon', border="black")  # Standard normal
            
            hist(r_unifgenFx(1000, qfunction = qunif), main = 'Uniform', xlab = 'p', col = 'green',border="black")
            par(oldPar) # Return to original graphical parameters
            ```
            
          iii. *(2 marks)* Generate a sample of 1000 pseudo-random observations from a 
              Student-t distribution on 3 degrees of freedom  generated using `r_unifgenFx()` (unchanged) and the quantile function of the Student-t.
              Plot a histogram (appropriately labelled) of the results.
              ```{r hist_t, eval = TRUE}
              set.seed(1234567)
              oldPar <- par(mfrow = c(1,2))
              hist(qt(r_unifgenFx(1000, qfunction = qunif), df=3), main = 'Student-t distribution', xlab = 'p', col = 'maroon')
              par(oldPar) # Return to original graphical parameters
              ```

    d. Consider the `quantile()` function in \R.
    
        i. *(2 marks)* Explain the values returned by `quantile(mtcars$qsec))`.  That is, what does `quantile()` do?
          A quantile, or percentile, tells you how much of your data lies below a certain value.
          ```{r, eval = TRUE}
            quantile(mtcars$qsec)
          ```
        For example, if we call quantile(mtcars$qsec), it returns that the first 25% values are in (14.5000, 16.8925), the second 25% of the values are in (16.8925, 17.7100). The third 25% values are in (17.7100, 18.9000) and the last 25% values are in (18.9000,22.9000)
        
        ii. *(2 marks)* Show how `quantile()` could be used to generate 1000 observations from the estimated distribution of `mtcars$qsec`.
```{r, eval = TRUE}
  estimated_qsec <- quantile(mtcars$qsec, probs = runif(1000, min = 0, max =1))
  
```
        iii. *(2 marks)* Would this work for `mtcars$cyl`? Why? Or, why not?
          No. The reason is the values if mtcars$cyl can be 4, 6, and 8 only. We cannot divide them into 4 quantiles.
        iv. *(4 marks)* Draw side by side (nicely labelled) histograms of `mtcars$qsec` and a sample of 1000 observations drawn from the estimated distribution of `mtcars$qsec`.  Comment on how these compare.
          ```{r, eval = TRUE}
          set.seed(1234567)
            oldPar <- par(mfrow = c(1,2))
            hist(mtcars$qsec,main = 'quarter-mile seconds', xlab = 'Seconds', col = 'yellow')
            hist(estimated_qsec, main = 'Estimated seconds', xlab = 'Seconds', col = 'maroon')
            par(oldPar) 
          ```
          
          I think the estimation is very close to true.
        
        v. *(3 marks)* Draw a (nicely labelled) `qqplot()` comparing the above two sets of observations.  What do you conclude about their empirical distributions?  Why?
          ```{r, eval = TRUE}
            qqplot(mtcars$qsec, estimated_qsec, main = 'qqplot, quarter-mile seconds and estimated distribution', xlab = 'real qsec (s)', ylab = 'estimation(s)')
          ```
          The estimation is percise because the y-intercept of the plot is close to 0 and the slope is close to 1.
    
        vi. *(4 marks)* Suppose interest lay in producing a bootstrap distribution for some estimator $\bigwig{\theta}$.  Instead of bootstrapping, how might `quantile()` be used?  Which would you recommend and why?
          We can further use the quantile function for each 25% of the data. By making 4x4=16 quantiles, the result could be more close to th real data.
          Bootstraping is more complex and the result may depend on the representative sample.